---
title: 3 game losing streak
author: Jay
date: '2018-01-30'
slug: 3-game-losing-streak
categories: [UNC]
tags: []
draft: true
---

Here we go again. 3 Ls in a row. When was the last time we had this kind of losing skid? 

```{r, include = FALSE}
suppressPackageStartupMessages(library(uncmbb))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(jutilr)) # dev

#losing_streaks <- function(df, n){ # n: number of losing streak
#aa <- df %>% group_by(Season) %>%
#            mutate(game_order = row_number()) %>%
#            mutate(result_enum = case_when(Result == "W" ~ 1,
#                                           Result == "L" ~ -1)) %>%
#            mutate(result_enum_cumulative = cumsum(result_enum))
           
# dynamically create new fields for next match result
#for (future in 1:(n+1)){ # need n+1 for first win after n-game losing streak
#  rname <- paste0("Result_", future) 
#  a <- aa %>% mutate(!!rname := lead(Result, future))
#}

# start with a win
#b <- a %>% filter(Result == "W", game_order != 1)
# first losing game
# b <- a %>% filter(Result == "L")

# dynamically filter rows based on next match result (note not using dplyr::filter)
#for (future in 1:n){
#  rname <- paste0("Result_", future)
#  b <- b[b[rname] == "L", ]
#  b <- b %>% filter(!is.na(Season)) # unlike the first win == "W" case, need to make sure all three #results are "L"
#}

# first win after L streaks
#rname <- paste0("Result_", n+1)
#b <- b[b[rname] == "W" | is.na(b[rname]), ] # use is.na to avoid season's last game 
#b
#head(b)
# now for seasons with first 3 games of the season all being Ls
#aa %>% filter((game_order == 3, result_enum_cumulative == -3), 
#              (game_order == 4, result_enum_cumulative == -2)) %>% head
#}

#streak <- 3
#streak_3Ls <- losing_streaks(unc, streak) %>%
#              count(Season)

#tail(streak_3Ls, 5) %>% kable("html", caption = "Last 5 seasons with 3 game losing streaks") %>%
#                        kable_styling()
```

So the last time there was a 3-game losing streak was back in 2014, and to my surprise,  

```{r, include = FALSE}
#streak <- 3
#a <- losing_streaks(unc, streak) %>% 
#     filter(Season == 2014) #%>%
#     #select(Season, Game_Date, starts_with("Result"))
#a
#unc %>% filter(Game_Date >= a$Game_Date) %>%
#        head(5) %>%
#        kable("html", caption = "2016 season 3 game losing streak records") %>%
#        kable_styling()
```

I got curious as to what the worst losing streak was since 1950 season. To do that I'm looking at the number of seasons with losing streaks up to certain nuber. 

```{r, include = FALSE}
#target_streaks <- 8 # code breaks when n = 9
#streaks <- vector("list", length = target_streaks)
#for (streak in 1:target_streaks){
#streaks[[streak]] <-  losing_streaks(unc, streak) %>% count(Season)
#}

#names(streaks) <- 1:target_streaks

#a <- tibble(streak = 1:target_streaks, seasons = streaks %>% map_int(function(x) nrow(x)), perc = round(seasons/67, 2)) 

#a %>% ggplot(aes(x = streak, y = perc)) +
#      geom_bar(stat = "identity") +
#      geom_text(aes(label = seasons), vjust = -0.1) +
#      theme_bw() +
#      scale_y_continuous(label = scales::percent) +
#      labs(title = "Number of seasons with losing streaks", x = "Losing streak", y = "Number of seasons")
            
```

So as seen earlier, one in every four seasons, we've seen at least one occasion of 3-game losing streaks. What gets me is that there's been seasons in which we had 4+ game losing streaks! Let's see what those seasons were and what were those games for those seasons with at least one 6-game losing streaks. 

```{r, include = FALSE}
#losing_streaks(unc, 6) %>%
#        kable("html") %>%
#        kable_styling()

#losing_streaks(unc, 7) %>%
#        kable("html") %>%
#        kable_styling()

#losing_streaks(unc, 8) %>%
#        kable("html") %>%
#        kable_styling()
```